### 2013-01-29 Docs/Thesis/background.tex
Revisions for MTF

The goal of FRP is to provide compositional and high-level abstractions for
creating reactive programs. The key abstractions of FRP are behaviors or
signals\footnote{Behaviors are generally referred to as {\em signals} in
signal-function literature. This is unfortunate, since a {\em signal function}
may operate on events, signals, or some combination of the two.}, which are
time-dependent values defined at every point in continuous time, and events,
which are values defined at countably many points in time. An FRP system will
provide combinators to manipulate events and signals and to react to events by
replacing a portion of the running program in response to an event. Behaviors
and events, or some abstraction based on them, will be first class, in keeping
with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable, but this has proven to be the main
challenge in implementing FRP.

The two general approaches to FRP are ``classic'' FRP, where behaviors and signals are first-class and reactive objects, and ``signal-function'' FRP,
where transformers of signals and events are first-class and reactive objects.

>>>>

The goal of FRP is to provide compositional and high-level abstractions for
creating reactive programs. The key abstractions of FRP are behaviors or
signals\footnote{Behaviors are generally referred to as {\em signals} in
signal-function literature. This is unfortunate, since a {\em signal function}
may operate on events, behaviors, or some combination of the two.}, which are
time-dependent values defined at every point in continuous time, and events,
which are values defined at countably many points in time. An FRP system will
provide combinators to manipulate events and behaviors and to react to events by
replacing a portion of the running program in response to an event. Behaviors
and events, or some abstraction based on them, will be first class, in keeping
with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable, but this has proven to be the main
challenge in implementing FRP.

The two general approaches to FRP are ``classic'' FRP, where behaviors and
events are first-class and reactive objects, and ``signal-function'' FRP, where
transformers of signals and events are first-class and reactive objects.

### 2013-01-29 Docs/Thesis/comparison.tex
Revisions for MTF

\section{Evaluation Application}
\label{section:Evaluation_and_Comparisons-Evaluation_Application}

The application of choice is a learning switch controller for OpenFlow~\cite{OpenflowSpec}
switches. this application must handle input events triggered by switches
signalling that they have received a packet for which they have no handling
rules, and produce output events which install handling rules for dealing with
these packets.

The reactive algorithm is simple. A table from machine addresses to switch ports
is maintained. This table is updated in two ways. When a packet is sent to the
controller, meta-data for the packet indicates which port it arrived on. The
source address of the packet can then be mapped to that port in the table.
Once this occurs, rules are installed for each packet with a destination address
matching the source, along with reciprocal rules.

Each mapping in the table has a timestamp. After a certain amount of time, the
table entry is deleted. The table entry is also deleted if a packet is received
with the same source address and a different port. When an entry is deleted,
the switch is signaled to remove all corresponding rules.

In both Yampa and TimeFlies, this algorithm is implemented by using an accumulator,
which applies functions carried by event occurrences to a state value. It stores
the new state and outputs new event occurrences based on the output of the function.

The state stored is the table (implemented using the {\tt Data.Map} module in the
Haskell {\tt containers} library) and the output event occurrences carry messages
to be sent to the switch. The events are sourced internally to the learning
switch signal function (using the {\tt after} combinator in TimeFlies and the {\tt repeatedly}
function in Yampa) to delete expired rules, and externally, by transforming messages
from the switch to the controller into accumulator functions.

@@

The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
amount to time to pause between time and signal updates was passed as a parameter to
both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
event, waiting for a rule response, and then repeating. The output of the benchmark
is the number of times this occurs in a second. This is thus a very good test of
event latency in a reactive system.

@@

\begin{figure}
\epsfig{file=graph.eps,angle=-90,width=\linewidth}
\hrule
\caption{Comparison of timeflies vs. Yampa implementing an OpenFlow learning switch controller.}
\label{fig:timeflies-yampa-comparison}
\end{figure}

There is some correlation between the sampling rate and the response rate
for TimeFlies, which is due to the continued wrapping of a replacement signal
function by {\tt switch} until the next time step (see Section~\ref{subsection:Implementation-Signal_Functions-Implementation_of_Signal_Function_Combinators}).
Nevertheless, TimeFlies outperforms Yampa in event response by several orders of magnitude,
even at a very high sampling rate. This is due to two factors. First, TimeFlies can react
to individual events without evaluating the whole signal function. Second, the
interface of time-flies permits us to connect its evaluation directly to GHC's event
manager, while the design of Yampa requires us to use threads and synchronized
communication\footnote{Yampa does include a step-wise evaluation function, but
it still couples event evaluation firmly to time steps, and is not well
documented.}.

>>>>

@@

The example application from Chapter~\ref{chapter:Example_Application}, an
OpenFlow controller, was also implemented in Yampa. A slightly-modified
TimeFlies implementation, which used a polling loop rather than the GHC event
system to obtain events, but still "pushed" events to TimeFlies once they were
acquired, was also implemented.

These threee controllers were benchmarked using the {\tt cbench}
program~\cite{cbench}. The amount to time to pause between time and signal
updates (the inverse of the sampling rate for signals) was passed as a parameter
to each program. The {\tt cbench} benchmark operates by sending a simulated
packet-in event, waiting for a rule response, and then repeating. The output of
the benchmark is the number of times this occurs in a second. This is thus a
test of event latency in a reactive system.

@@

\begin{figure}
\includegraphics{graph}
\hrule
\caption{Comparison of timeflies vs. Yampa implementing an OpenFlow learning switch controller.}
\label{fig:timeflies-yampa-comparison}
\end{figure}

There is some correlation between the sampling rate and the response rate
for TimeFlies, which is due to the continued wrapping of a replacement signal
function by {\tt switch} until the next time step (see Section~\ref{subsection:Implementation-Signal_Functions-Implementation_of_Signal_Function_Combinators}).
Nevertheless, TimeFlies outperforms Yampa in event response by several orders of magnitude,
even at a very high sampling rate. This is due to two factors. First, TimeFlies can react
to individual events without evaluating the whole signal function. Second, in
the implementation described in Chapter~\ref{chapter:Example_Application}, the
interface of TimeFlies permits us to connect its evaluation directly to GHC's
event manager, while the design of Yampa requires us to use threads and
synchronized communication\footnote{Yampa does include a step-wise evaluation
function, but it still couples event evaluation firmly to time steps, and is not well
documented.}. Even the TimeFlies implementation using polling IO outperforms the
Yampa version, due to the ability to evaluate multiple events per time step.

### 2013-01-29 Docs/Thesis/conclusions-and-further-work.tex
Revisions for MTF

\section{Conclusions}
\label{section:Conclusions_and_Further_Work-Conclusions}
The TimeFlies system is a fully-implemented and -documented FRP library which
may be extended with utility functions and further optimized. Its performance in
responding to events is demonstrated to be superior to that of the predominant
pull-based signal-function FRP library.

@@

Many classes of reactive application would benefit from a ``dynamic collections'' combinator
similar to {\tt pSwitch} in Yampa. Such a combinator allows a collection of signal functions
to be evaluated as one signal function, with addition or removal of signal functions instead
of the total replacement given by the {\tt switch} combinator. This is useful, for instance,
when simulating objects for games or computer models, as the behavior of each object can be
modeled as a signal function, and these signal functions can be added to and removed from the
collection.

>>>>

\section{Conclusions}
\label{section:Conclusions_and_Further_Work-Conclusions}
The TimeFlies system is a fully-implemented and fully-documented FRP library which
may be extended with utility functions and further optimized. Its performance in
responding to events is demonstrated to be superior to that of the predominant
pull-based signal-function FRP library.

@@

Many classes of reactive application would benefit from a ``dynamic collections'' combinator
similar to {\tt pSwitch} in Yampa. Such a combinator allows a collection of signal functions
to be evaluated as one signal function, with addition or removal of signal functions instead
of the total replacement given by the {\tt switch} combinator. This is useful, for instance,
when simulating objects for games or computer models, as the behavior of each object can be
modeled as a signal function, and these signal functions can be added to and removed from the
collection. An attempt was made to add such a function to TimeFlies. Currently,
it is not clear what the type of such a combinator would be in a system
employing signal vectors, or how the "routing" of inputs to individual signal
functions in the collection would be specified.

### 2013-01-29 Docs/Thesis/introduction.tex
Revisions for MTF

Functional reactive programming (FRP) is a promising class of abstractions for
writing interactive programs in functional languages. The FRP abstractions are
behaviors or signals, which are functions of continuous time, and events, which
are sequences of time-value pairs. These abstractions translate desirable
properties of the underlying functional languages, such as higher-order
functions and referential transparency, to reactive constructs, generally
without modifications to the underlying language. This permits the compositional
construction of reactive programs in purely functional languages.

Functional reactive programming was introduced, though not quite by name, with Fran~\cite{Elliott1997}, a compositional system for writing
reactive animations. From the start, two key challenges emerged: the enforcement of {\em causality}\footnote{Causality is the property 
that a value in an FRP program depends only on present or past values. Obviously, a program which violates this property is impossible to
evaluate, but in some FRP systems such programs are unfortunately quite easy to write down.} in FRP programs, and the efficient
evaluation of FRP programs.

The first challenge was addressed by representing FRP programs, not as
compositions of signals and events, but as compositions of transformers of
behaviors and events called {\em signal functions}. By not permitting direct
access to behaviors or events, but representing them implicitly, signal
functions prohibit accessing past or future time values, only allowing access to
values in the closure of the signal function and current input values. Signal
function FRP programs are written by composing signal functions, and are
evaluated using a function which provides input to the signal function and acts
on the output values. This evaluation function is generally elided in existing
literature on signal functions, but we provide a documented and improved
interface for evaluating signal functions. A further advantage of
signal-function FRP programs is that they are more readily composed,
since additional signal functions may be composed on the input or output side of
a signal function, rather than simply the output side as in classic
(non-signal-function) FRP systems.

@@

To address both challenges in a FRP implementation, we combine the approaches
and demonstrate a push-based signal-function FRP system. The current
implementation approach for signal-function FRP is inherently
pull-based~\cite{Nilsson2002}, but recent work on N-ary
FRP~\cite{Sculthorpe2011}, a variant of signal-function FRP which enumerates
inputs and distinguishes between events and behaviors, provides a representation
which suggests a method of push-based evaluation. In previous implementations of
FRP, events were simply option-valued signals. This approach has the advantage
of being simple to implement. However, it does not have an obvious semantics for
event merging. Further, it necessitates pull-based evaluation, since there is no
way to separate the events which may be pushed from the truly continuous values
which must be repeatedly sampled.

>>>>

Functional reactive programming (FRP) is a promising class of abstractions for
writing interactive programs in functional languages. The FRP abstractions are
{\em behaviors} (also termed {\em signals}), which are functions of continuous
time, and {\em events}, which are sequences of time-value pairs. These
abstractions translate desirable properties of the underlying functional
languages, such as higher-order functions and referential transparency, to
reactive constructs, generally without modifications to the underlying language.
This permits the compositional construction of reactive programs in purely
functional languages.

Functional reactive programming was introduced, though not quite by name, with
Fran~\cite{Elliott1997}, a compositional system for writingm reactive
animations. From the start, two key challenges emerged: the enforcement of
{\em causality}\footnote{Causality is the property that a value in an FRP
program depends only on present and past values. Obviously, a program which
violates this property is impossible to evaluate, but in some FRP systems such
programs are unfortunately quite easy to write down.} in FRP programs, and the
efficient evaluation of FRP programs.

The first challenge was addressed by representing FRP programs, not as
compositions of behaviors and events, but as compositions of transformers of
behaviors and events called {\em signal functions}. By not permitting direct
access to behaviors or events, but representing them implicitly, signal
functions prohibit accessing past or future time values, only allowing access to
values in the closure of the signal function and current input values. Signal
function FRP programs are written by composing signal functions, and are
evaluated using a function which provides input to the signal function and acts
on the output values. This evaluation function is generally elided in existing
literature on signal functions, but we provide a documented and improved
interface for evaluating signal functions. A further advantage of
signal-function FRP programs is that they are more readily composed,
since additional signal functions may be composed on the input or output side of
a signal function, rather than simply the output side as in classic
(non-signal-function) FRP systems.

@@

To address both challenges in a FRP implementation, we combine the approaches
and demonstrate a push-pull signal-function FRP system. The current
implementation approach for signal-function FRP is inherently
pull-based~\cite{Nilsson2002}, but recent work on N-ary
FRP~\cite{Sculthorpe2011}, a variant of signal-function FRP which enumerates
inputs and distinguishes between events and behaviors, provides a representation
which suggests a method of push-based evaluation. In previous implementations of
signal-function FRP, events were simply option-valued signals. This approach has
the advantage of being simple to implement. However, it does not have an obvious
semantics for event merging. Further, it necessitates pull-based evaluation,
since there is no way to separate the events which may be pushed from the truly
continuous values which must be repeatedly sampled.

### 2013-01-29 Docs/Thesis/system_design_and_interface.tex
Revisions for MTF

The evaluation monad is a monad transformer. This permits it to be used in
conjunction with the {\tt IO} monad (or any other monad) to describe how input
is to be obtained for the signal function being evaluated, and how outputs are
to be handled.

@@

The type of the evaluation monad must track the input and output type of the
signal function. The monad's context stores a mapping from outputs to handling
actions. There are thus four type parameters to the monad's type
constructor: the input signal vector, the output signal vector, the type of the
underlying monad, and the monadic type parameter. The type is shown in
Figure~\ref{figure:evaluation_monad_types}.

>>>>

In TimeFlies, the evaluation monad is a monad transformer. This permits it to be
used in conjunction with the {\tt IO} monad (or any other monad) to describe how
input is to be obtained for the signal function being evaluated, and how outputs
are to be handled.

@@

The type of the evaluation monad must track the input and output type of the
signal function. The monad is a state monad, and its state is a record with
fields for a mapping from outputs to handling actions, the current state of the
signal function, the last time the signal function was sampled, and the
currently-unsampled input. There are thus four type parameters to the monad's
type constructor: the input signal vector (to type the signal function and
unsampled input), the output signal vector (to type the signal function and
output handlers), the type of the underlying monad, and the monadic type
parameter. The type is shown in Figure~\ref{figure:evaluation_monad_types}.
